On the power of conditional independence testing \\ under model-X

### 2022-04-26 arxiv-v4/response_to_reviews.tex

\subsection{Major comments}

\begin{enumerate}
	\item \textsl{I do find the different sections and aims of the paper a bit disconnected— specifically I’m not sure if (1) is well-connected with (2) and (3). (1) is to do with the power of the CRT test. On the other hand, (2) is about a different test that works with different assumptions, though there is an asymptotically equivalent CRT test, and (3) studies the power of this test. The motivation behind (2) seems to be concerns about the validity of the CRT assumptions (which require full knowledge of the conditional distribution $X|Z$), but I don’t see how this idea of ‘robust- ness’ really relates closely to (1). One option could be to remove or de-emphasize (1)—in my opinion it is nice to have such a result written down somewhere, but it is unsurprising.}\\

	We thank the referee for pointing out the somewhat disconnected nature of the submitted manuscript, especially the transition from Section 2 to Section 3. We agree, and we have substantially edited the exposition in Section 3 for a smoother flow. The high-level logic in the revised manuscript is that Section 2 studies the power and optimality of the CRT in finite samples against point alternatives while Sections 3 and 4 study the power of the CRT in an asymptotic setting against semiparametric alternatives. Section 3 prepares the groundwork by establishing the asymptotic equivalence between the distilled CRT and the simpler-to-analyze MX(2) $F$-test (loosening the assumption from MX to MX(2) along the way), while Section 4 establishes the power of the MX(2) $F$-test and therefore the distilled CRT. Ultimately, the central goal of the manuscript is to study the power and optimality of the CRT, and the MX(2) $F$-test is mostly a means to that end---though we argue it is of independent interest as well. This logic is conveyed in the opening paragraph of Section 3 (pp. 8-9):

	\begin{quote}
		``In Section 2, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\bm Y| \bm X, \bm Z}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this machine learning procedure. Specifically, we consider  the power of a specific instance of the CRT against a set of semiparametric alternatives (Section 3.1). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section 3.2), which is of independent interest due to its closed form and weaker assumptions (Section 3.3). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section 3.4) and put this section's results into perspective (Section 3.5) before moving on to stating the desired power results in the next section (Section 4).''
	\end{quote}

### 2022-04-26 arxiv-v4/crtpower.tex
Changed ML to machine learning

The CI testing problem under MX has benefited from several methodological innovations, but fundamental questions regarding power and optimality have received less attention. Therefore, in this paper we address the following two primary questions:
\begin{enumerate}
	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
	\item[Q2.] What is the precise connection between the performance of the machine learning (ML) algorithm and the power of the resulting MX method?
\end{enumerate}
To the best of our knowledge, Q1 has not been considered before, while Q2 has only been indirectly addressed in the context of lasso-based knockoffs \cite{Weinstein2017, Liu2019, Fan2020, Weinstein2020} and CRT \cite{Wang2020b, Celentano2020}. The present paper complements these existing works by considering arbitrary ML methods. We summarize our findings next.

@@

While the composite nonparametric alternative of the CI problem~\eqref{conditional-independence} suggests that we cannot expect to find a uniformly most powerful test, we may still ask what is the most powerful test against a point alternative. We show that any level $\alpha$ test must also have level $\alpha$ conditionally on $(\sry, \srz)$, which allows us to reduce the composite null to a point null. We can therefore apply the Neyman-Pearson lemma to show (Section~\ref{sec:power}) that the optimal test against a point alternative $\mathcal L$ with $\mathcal L(\pry|\prx,\prz) = \bar f_{\pry|\prx,\prz}$ is the CRT based on the likelihood test statistic:
\begin{equation}
	T^{\text{opt}}(\srx; \sry, \srz) \equiv \prod_{i = 1}^n \bar f(\sry_i|\srx_i, \srz_i).
\end{equation}
The same statistic yields the most powerful one-bit $p$-values for MX knockoffs (Section~\ref{sec:knockoffs}). Despite the simplicity of the final answer, this result has not been derived before, and appears central to the design of good (powerful) test statistics. Since the model for $Y|X,Z$ is unknown, this result provides our first theoretical indication of the usefulness of ML models to learn this distribution (Q2). A2 below gives a more quantitative answer to Q2.

\paragraph{A2: The prediction error of the ML method impacts the asymptotic efficiency of the CRT but not its consistency.}

It has been widely observed that the better the ML method approximates $\pry|\prx,\prz$, the higher power the MX method will have. We put this empirical knowledge on a theoretical foundation by expressing the asymptotic power of the CRT in terms of the prediction error of the underlying ML method (Section~\ref{sec:asymptotic-power}). In particular, we consider semiparametric alternatives of the form
\begin{equation}
	H_1: \mathcal L(\pry|\prx,\prz) = N(\prx^T\beta + g(\prz),\sigma^2).
	\label{parametric-alternative-intro}
\end{equation}
We analyze the power of a CRT variant that employs a separately trained estimator $\widehat g$ in an asymptotic regime where $d = \text{dim}(\prx)$ remains fixed while $p = \text{dim}(\prz)$ grows arbitrarily with the sample size $n$. We find that this test is consistent no matter what $\widehat g$ is used, while its asymptotic power against local alternatives $\beta_n = h/\sqrt{n}$ depends on the limiting mean-squared prediction error of $\widehat g$ (denoted $\mathcal E^2$) and the limiting expected variance $\mathbb E[\text{Var}[\prx | \prz]]$ (denoted $s^2$). For example, if $d = 1$, the
\begin{equation*}
	\begin{split}
		\text{CRT power converges to that of normal location test under alternative } N\smash{\left(\frac{hs}{\sqrt{\sigma^2 + \mathcal E^2}},1\right)}.
	\end{split}
\end{equation*}
This represents the first explicit quantification of the impact of ML prediction error on the power of an MX method. \\

@@

In Section~\ref{sec:power}, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\pry|\prx,\prz}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this ML procedure. Specifically, we consider  the power of a specific instance of the CRT (the \textit{distilled CRT} \cite{Liu2020})  against a set of semiparametric alternatives (Section~\ref{sec:asymptotic-setup}). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section~\ref{sec:second-order-approx}), which is of independent interest due to its closed form and weaker assumptions (Section~\ref{sec:mx2-f-test}). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section~\ref{sec:simulations}) and put this section's results into perspective (Section~\ref{sec:comparison-to-existing-results-3}) before moving on to stating the desired power results in the next section (Section~\ref{sec:asymptotic-power}).

@@

In this paper, we gave some answers to the theoretical questions posed in the introduction. We presented the first finite-sample optimality results in the MX framework and explicitly quantified how the performance of the underlying ML procedure impacts the asymptotic power of the CRT. Along the way, we exhibited a weakened form of the MX assumption and a resampling-free methodology valid under only this assumption.

@@

The MX framework is just one setting where black-box prediction methods have been recently employed for the purpose of more powerful statistical inference. Other examples include conformal prediction \cite{FoygelBarber2019}, classification-based two-sample testing \cite{Kim2020} and data-carving based multiple testing \cite{lei2016adapt}. These methods employ ML algorithms to create powerful test statistics, calibrating them for valid inference with no assumptions about the method used. However, the more accurate the learned model, the more powerful the inference. Our finite-sample and asymptotic power results explicitly tie the error of the learning algorithm to the power of the test, and thus put this common intuition on a quantitative foundation and may thus help inform the choice and design of ML methods used for inferential goals.

>>>>

The CI testing problem under MX has benefited from several methodological innovations, but fundamental questions regarding power and optimality have received less attention. Therefore, in this paper we address the following two primary questions:
\begin{enumerate}
	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
	\item[Q2.] What is the precise connection between the performance of the machine learning algorithm and the power of the resulting MX method?
\end{enumerate}
To the best of our knowledge, Q1 has not been considered before, while Q2 has only been indirectly addressed in the context of lasso-based knockoffs \cite{Weinstein2017, Liu2019, Fan2020, Weinstein2020} and CRT \cite{Wang2020b, Celentano2020}. The present paper complements these existing works by considering arbitrary machine learning methods. We summarize our findings next.

@@

While the composite nonparametric alternative of the CI problem~\eqref{conditional-independence} suggests that we cannot expect to find a uniformly most powerful test, we may still ask what is the most powerful test against a point alternative. We show that any level $\alpha$ test must also have level $\alpha$ conditionally on $(\sry, \srz)$, which allows us to reduce the composite null to a point null. We can therefore apply the Neyman-Pearson lemma to show (Section~\ref{sec:power}) that the optimal test against a point alternative $\mathcal L$ with $\mathcal L(\pry|\prx,\prz) = \bar f_{\pry|\prx,\prz}$ is the CRT based on the likelihood test statistic:
\begin{equation}
	T^{\text{opt}}(\srx; \sry, \srz) \equiv \prod_{i = 1}^n \bar f(\sry_i|\srx_i, \srz_i).
\end{equation}
The same statistic yields the most powerful one-bit $p$-values for MX knockoffs (Section~\ref{sec:knockoffs}). Despite the simplicity of the final answer, this result has not been derived before, and appears central to the design of good (powerful) test statistics. Since the model for $Y|X,Z$ is unknown, this result provides our first theoretical indication of the usefulness of machine learning models to learn this distribution (Q2). A2 below gives a more quantitative answer to Q2.

\paragraph{A2: The prediction error of the machine learning method impacts the asymptotic efficiency of the CRT but not its consistency.}

It has been widely observed that the better the machine learning method approximates $\pry|\prx,\prz$, the higher power the MX method will have. We put this empirical knowledge on a theoretical foundation by expressing the asymptotic power of the CRT in terms of the prediction error of the underlying machine learning method (Section~\ref{sec:asymptotic-power}). In particular, we consider semiparametric alternatives of the form
\begin{equation}
	H_1: \mathcal L(\pry|\prx,\prz) = N(\prx^T\beta + g(\prz),\sigma^2).
	\label{parametric-alternative-intro}
\end{equation}
We analyze the power of a CRT variant that employs a separately trained estimator $\widehat g$ in an asymptotic regime where $d = \text{dim}(\prx)$ remains fixed while $p = \text{dim}(\prz)$ grows arbitrarily with the sample size $n$. We find that this test is consistent no matter what $\widehat g$ is used, while its asymptotic power against local alternatives $\beta_n = h/\sqrt{n}$ depends on the limiting mean-squared prediction error of $\widehat g$ (denoted $\mathcal E^2$) and the limiting expected variance $\mathbb E[\text{Var}[\prx | \prz]]$ (denoted $s^2$). For example, if $d = 1$, the
\begin{equation*}
	\begin{split}
		\text{CRT power converges to that of normal location test under alternative } N\smash{\left(\frac{hs}{\sqrt{\sigma^2 + \mathcal E^2}},1\right)}.
	\end{split}
\end{equation*}
This represents the first explicit quantification of the impact of machine learning prediction error on the power of an MX method. \\

@@

In Section~\ref{sec:power}, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\pry|\prx,\prz}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this machine learning procedure. Specifically, we consider  the power of a specific instance of the CRT (the \textit{distilled CRT} \cite{Liu2020})  against a set of semiparametric alternatives (Section~\ref{sec:asymptotic-setup}). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section~\ref{sec:second-order-approx}), which is of independent interest due to its closed form and weaker assumptions (Section~\ref{sec:mx2-f-test}). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section~\ref{sec:simulations}) and put this section's results into perspective (Section~\ref{sec:comparison-to-existing-results-3}) before moving on to stating the desired power results in the next section (Section~\ref{sec:asymptotic-power}).

@@

In this paper, we gave some answers to the theoretical questions posed in the introduction. We presented the first finite-sample optimality results in the MX framework and explicitly quantified how the performance of the underlying machine learning procedure impacts the asymptotic power of the CRT. Along the way, we exhibited a weakened form of the MX assumption and a resampling-free methodology valid under only this assumption.

@@

The MX framework is just one setting where black-box prediction methods have been recently employed for the purpose of more powerful statistical inference. Other examples include conformal prediction \cite{FoygelBarber2019}, classification-based two-sample testing \cite{Kim2020} and data-carving based multiple testing \cite{lei2016adapt}. These methods employ machine learning algorithms to create powerful test statistics, calibrating them for valid inference with no assumptions about the method used. However, the more accurate the learned model, the more powerful the inference. Our finite-sample and asymptotic power results explicitly tie the error of the learning algorithm to the power of the test, and thus put this common intuition on a quantitative foundation and may thus help inform the choice and design of machine learning methods used for inferential goals.
